Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
Fixes
Compact

\paragraph{Both $ X $ and $ Y $ are continuous: }
When both $ X $ and $ Y $ are continuous, Pearson's correlation test can be
used and Pearson's correlation coefficient can be used at the effect size. When
$ \bm{Z} = \emptyset $, the correlation coefficient is defined as:

\begin{equation}
	r_{X, Y} = \frac{\mathrm{cov}(X, Y)}{\sigma_X \sigma_Y}
\end{equation}

When $ \bm{Z} \neq \emptyset $, partial correlation coefficient can be used.
This is estimated by fitting two regression models $ E_X: X \sim \bm{Z} $ and $
E_Y: Y \sim \bm{Z} $, calculating the residuals $ R_X = X - E_X(\bm{Z}) $ and $
R_Y = Y - E_Y(\bm{Y}) $, and then computing the correlation coefficient between
the residuals:

\begin{equation}
	r_{X, Y \rvert \bm{Z}} = r_{R_X, R_Y}
\end{equation}

@@

Polyserial (when one is ordinal and the other is continuous) and Polychoric
(when both are ordinal) correlation have been used to estimate the covariance
matrix between them \citep{Poon1987}. Both methods make the assumption that the
observed ordinal variable is a result of thresholding a latent normally
distributed continuous variable. Under this assumption, the methods then try to
estimate the covariance matrix while maximizing the likelihood of the dataset.
Using the estimated covariance matrix, $ \Sigma $ we can perform the CI test and compute 
the Pearson's correlation coefficient as the effect size:

\begin{equation}
	r_{X, Y \rvert \bm{Z} = \emptyset} = \frac{\Sigma_{XY}}{\sqrt{\Sigma_{XX} \Sigma_{YY}}}; \;\;\; \; r_{X, Y \rvert \bm{Z}} = - \frac{\Sigma^{-1}_{XY}}{\sqrt{\Sigma^{-1}_{XX} \Sigma^{-1}_{YY}}}
\end{equation}

@@

In the case when we have a combination of ordinal and categorical variables, we
can use a residualization based CI test \citep{Ankan2023}. This test gives a
chi-square distributed test statistic, that can be used for computing an effect
size such as the Root Mean Squared Error of Approximation (RMSEA). Given a
chi-squared test statistic, $ \chi^2 $, with $ \textrm{df} $ degrees of
freedom, the RMSEA is defined as:

\begin{equation}
	\textrm{RMSEA}_{X, Y \rvert \bm{Z}} = \sqrt{\frac{\chi^2 - \textrm{df}}{\textrm{df} (N-1)}}
\end{equation}

where $ N $ is the sample size.

@@

\begin{equation}
	\begin{split}
		x &= \textrm{rand}([0, 1]) \\
		\mathrm{Expert}(\alpha) &= \begin{cases} 
			\mathrm{TrueReln}(X, Y),  & \textrm{if  } x <= \alpha \\
			\textrm{rand}(X \rightarrow Y, Y \leftarrow X, \textrm{None}) & \textrm{otherwise} \\
				\end{cases} \\
		\mathrm{TrueReln}(X, Y) &= \begin{cases}
				X \rightarrow Y, & \textrm{if } X \rightarrow Y \in G_{\textrm{true}} \\
				Y \rightarrow X, & \textrm{if } Y \rightarrow X \in G_{\textrm{true}} \\
					\textrm{None}, & \textrm{otherwise }
				  \end{cases}
	\end{split}
\end{equation}

>>>>

\paragraph{Both $ X $ and $ Y $ are continuous: }
When both $ X $ and $ Y $ are continuous, Pearson's correlation test can be
used and Pearson's correlation coefficient can be used at the effect size. When
$ \bm{Z} = \emptyset $, the correlation coefficient is defined as: $ r_{X, Y} =
\mathrm{cov}(X, Y) / (\sigma_X \sigma_Y) $. When $ \bm{Z} \neq \emptyset $,
partial correlation coefficient can be used. This is estimated by fitting two
regression models $ E_X: X \sim \bm{Z} $ and $ E_Y: Y \sim \bm{Z} $,
calculating the residuals $ R_X = X - E_X(\bm{Z}) $ and $ R_Y = Y - E_Y(\bm{Y})
$, and then computing the correlation coefficient between the residuals: $r_{X,
Y \rvert \bm{Z}} = r_{R_X, R_Y}$

@@

Polyserial (when one is ordinal and the other is continuous) and Polychoric
(when both are ordinal) correlation have been used to estimate the covariance
matrix between them \citep{Poon1987}. Both methods make the assumption that the
observed ordinal variable is a result of thresholding a latent normally
distributed continuous variable. Under this assumption, the methods then try to
estimate the covariance matrix while maximizing the likelihood of the dataset.
Using the estimated covariance matrix, $ \Sigma $ we can perform the CI test
and compute the Pearson's correlation coefficient as the effect size. When 
$\bm{Z} = \emptyset $, $ r_{X, Y} = \Sigma_{XY} / (\sqrt{\Sigma_{XX} \Sigma_{YY}}) $, 
	and when $ \bm{Z} \ne \emptyset $,
	$ r_{X, Y \rvert \bm{Z}} = - \Sigma^{-1}_{XY}/ (\sqrt{\Sigma^{-1}_{XX} \Sigma^{-1}_{YY}}) $.

@@

In the case when we have a combination of ordinal and categorical variables, we
can use a residualization based CI test \citep{Ankan2023}. This test gives a
chi-square distributed test statistic, that can be used for computing an effect
size such as the Root Mean Squared Error of Approximation (RMSEA). Given a
chi-squared test statistic, $ \chi^2 $, with $ \textrm{df} $ degrees of
freedom, the RMSEA is defined as: $ \textrm{RMSEA}_{X, Y \rvert \bm{Z}} =
\sqrt{\chi^2 - \textrm{df}/ (\textrm{df} (N-1))} $, where $ N $ is the sample
size.

@@

\begin{equation}
	\begin{split}
		x &= \textrm{rand}([0, 1]) \\
		\mathrm{Expert}(\alpha) &= \begin{cases} 
			\mathcal{O}(X, Y),  & \textrm{if  } x <= \alpha \\
			\textrm{rand}(X \rightarrow Y, Y \leftarrow X, \textrm{None}) & \textrm{otherwise} \\
				\end{cases} \\
			\mathcal{O}(X, Y) &= \begin{cases}
				X \rightarrow Y, & \textrm{if } X \rightarrow Y \in G_{\textrm{true}} \\
				Y \rightarrow X, & \textrm{if } Y \rightarrow X \in G_{\textrm{true}} \\
					\textrm{None}, & \textrm{otherwise }
				  \end{cases}
	\end{split}
\end{equation}
